---
title: "data_manipulation"
author: "Shivalika Chavan"
date: "2025-09-18"
output: github_document
---

```{r, message = FALSE}
library(tidyverse)
litters_df = 
  read_csv(
    "./data/FAS_litters.csv", 
    na = c("NA", ".", "")
    )
litters_df = 
  janitor::clean_names(litters_df)
pups_df = 
  read_csv(
    "./data/FAS_pups.csv", 
    na = c("NA", ".", ""), 
    skip = 3
    )
pups_df = 
  janitor::clean_names(pups_df)
```

## Select 
```{r collapse=TRUE}
select(litters_df, group, litter_number, gd0_weight, pups_born_alive)
select(litters_df, group:gd_of_birth)
select(litters_df, -pups_survive)
select(litters_df, GROUP = group, LiTtEr_NuMbEr = litter_number)
select(litters_df, starts_with("gd"))
select(litters_df, ends_with("birth"))
select(litters_df, contains("pups"))
select(litters_df, litter_number, pups_survive, everything()) #moves litter_number and pups to front of selection
relocate(litters_df, litter_number, pups_survive) #similar to everything()

# Learning Assessment
select(pups_df, litter_number, sex, pd_ears)
```

## Filter 
```{r collapse=TRUE}
filter(litters_df, gd_of_birth == 20)
filter(litters_df, pups_born_alive >= 2) # all litters have 3+ pups born alive
filter(litters_df, pups_survive != 4)
filter(litters_df, !(pups_survive == 4)) # same as line above
filter(litters_df, group %in% c("Con7", "Con8")) #useful for sets
filter(litters_df, group == "Con7" & gd_of_birth == 20)
```
It's common to omit missing observations. use `drop_na` for this from the `tidyr` package. 
```{r collapse=TRUE}
drop_na(litters_df) # drop rows with any value missing
# Learning Assessment

filter(pups_df, sex == 1)
filter(pups_df, sex == 2, pd_walk < 11)
```

## Mutate
```{r collapse=TRUE}
mutate(litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)

#Learning Assessment
mutate(
  pups_df, 
  pd_pivot_sub_7 = pd_pivot - 7,
  pd_sum = pd_ears + pd_eyes + pd_pivot + pd_pivot_sub_7 + pd_walk
  )
```

## Arrange
```{r collapse=TRUE}
head(arrange(litters_df, group, pups_born_alive), 10)
```

## />
super clunky options
```{r collapse=TRUE}
litters_df_raw = 
    read_csv("./data/FAS_litters.csv", na = c("NA", ".", ""))
litters_df_clean_names = janitor::clean_names(litters_df_raw)
litters_df_selected_cols = select(litters_df_clean_names, -pups_survive)
litters_df_with_vars = 
  mutate(
    litters_df_selected_cols, 
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group))
litters_df_with_vars_without_missing = 
  drop_na(litters_df_with_vars, wt_gain)
litters_df_with_vars_without_missing
```

less clunky second option
```{r collapse=TRUE}
litters_df_clean = 
  drop_na(
    mutate(
      select(
        janitor::clean_names(
          read_csv("./data/FAS_litters.csv", na = c("NA", ".", ""))
          ), 
      -pups_survive
      ),
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
    ),
  wt_gain
  )
litters_df_clean
```

Both give the same output but the code is not easy to read at all
Yay for piping!!! We <3 piping
```{r collapse=TRUE}
litters_df = 
  read_csv("./data/FAS_litters.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  select(-pups_survive) |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) |> 
  drop_na(wt_gain)
litters_df
```

Sometime you have to tell the data where to go in the pipe (never really needed in `tidyverse`)
```{r collapse=TRUE}
litters_df |>
  lm(wt_gain ~ pups_born_alive, data = _) |>
  broom::tidy()

# Learning Assessment
pups_df_LA =
  read_csv("./data/FAS_pups.csv", na = c("NA", ".", ""), skip = 3) |> 
  janitor::clean_names() |> 
  filter(sex == 1) |> 
  select(-pd_ears) |> 
  mutate(pd_pivot_more_than_7 = pd_pivot > 7)
pups_df_LA 
```



